---
title: "DATA606_FALL_2018_FINAL_PROJECT_PROPOSAL"
author: "John K. Hancock"
date: "October 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#import the required libraries
library(dplyr)
library(httr)
library(jsonlite)
library(tidyverse)  
library(rvest)    
library(stringr)   
library(rebus)
library(tm)
library(wordcloud)
library(viridisLite)
library(ggplot2)
library(stringr)
library(lubridate)
library(GENEAread)
library(Hmisc)
theme_update(plot.title = element_text(hjust = 0.5))
```


### Data Preparation
```{r}
mlb_df <- read.csv('DATA/MLB_PITCHING_STATS_1998_to_2017.csv')
head(mlb_df, 10)
```

```{r}
mlb_df[mlb_df$W >= 100, "Record"] <- "100 wins or Above"
mlb_df[mlb_df$W >= 82  & mlb_df$W < 100, "Record"] <- "Above .500 and Less than 100 Wins"
mlb_df[mlb_df$W < 82 & mlb_df$W > 50, "Record"] <- "Below .500 but above 50 wins"
mlb_df[mlb_df$W <= 50, "Record"] <- "Less than 50 wins"
mlb_df$Record <- as.factor(mlb_df$Record)

```

```{r}
mlb_df$League <- case_when(mlb_df$Team == 'Braves'~'National League',
 mlb_df$Team == 'Marlins'~'National League',
 mlb_df$Team == 'Mets'~'National League',
 mlb_df$Team == 'Phillies'~'National League',
 mlb_df$Team == 'Nationals'~'National League',
 mlb_df$Team == 'Cubs'~'National League',
 mlb_df$Team == 'Reds'~'National League',
 mlb_df$Team == 'Brewers'~'National League',
 mlb_df$Team == 'Pirates'~'National League',
 mlb_df$Team == 'Cardinals'~'National League',
 mlb_df$Team == 'iamondbacks'~'National League',
 mlb_df$Team == 'Rockies'~'National League',
 mlb_df$Team == 'Dodgers'~'National League',
 mlb_df$Team == 'Padres'~'National League',
 mlb_df$Team == 'Giants'~'National League',
 TRUE ~ as.character('American League'))
```




```{r}
rows <- as.numeric(row.names.data.frame(mlb_df[mlb_df$Team == 'Astros',])) 
rows <- rows[rows < 480]
mlb_df$League[ rows] <- "National League"
```


```{r}
mlb_df[mlb_df$Team == 'Astros', ]
```
### Research question 

Does a mlb team's key pitching statistics predict their win totals? If so, which pitching metric, ERA, Strikeouts, WHIP, or FIP provide the best predictor for wins? 


### Cases

The data above consists of each MLB team's wins, losses, ERA, SO, WHIP, and FIP for the past 20 MLB seasons from 1998 to 2017. The cutoff year of 1998 was chosen since that was the last year of MLB expansion. Thus, there are 600 cases.

### Data Collection

Data was collected from the website FanGraphs.com. For the purpose of this academic exercise, I purchased a year's membership to the site and I was able to download the datasets needed for this project.

### Type of study 

The data is observational.  Statistics are compiled from 30 teams over 20 years--1998 to 2018



### Data Source 

Fangraphs.com is website operated by FanGraphs, Inc. Fangraphs compiles historical statistical data for the entire history of Major League Baseball. In addition, it creates and records advanced baseball metrics outside of the established statistics. FanGraphs is well established as a chronicler and compiler of baseball statistics. It has parternership deals with ESPN and SB Nation. Link to website: [FanGraphs.com](https://www.fangraphs.com/) [Wikipedia](https://en.wikipedia.org/wiki/Fangraphs)



### Dependent Variable

There are two response variables. One is Wins which is numeric (quantative variable), and the other is Record which is qualitative.



### Independent Variable

The independent variables are ERA, Strikeouts("SO"), Walks, Hits to Innings Pitched ("WHIP), Fielded Independent Pitching ("FIP") which are all numeric.  The qualitative variable is League.


```{r}
describe(mlb_df$ERA)
```
```{r}
describe(mlb_df$SO)
```
```{r}
describe(mlb_df$WHIP)
```


```{r}
describe(mlb_df$FIP)
```

```{r}
table(mlb_df$Record, useNA='ifany')
```

```{r}
ggplot(mlb_df, aes(x=W)) + geom_histogram(bins=15,
                 col="blue", 
                 fill="orange") + labs(title="Histogram for Wins", align="center") +
                                  labs(x="Wins", y="Count")
```


```{r}
ggplot(mlb_df, aes(x=ERA)) + geom_histogram(bins=20,
                 col="white", 
                 fill="red") + labs(title="Histogram for ERA") +
                                  labs(x="Earned Run Average", y="Count")
```


```{r}
ggplot(mlb_df, aes(x=FIP)) + geom_histogram(bins=15,
                 col="blue", 
                 fill="orange") + labs(title="Histogram for Fielding Independent Pitching") +
                                    labs(x="FIP", y="Count")
```

