---
title: "DATA606_FALL_2018_FINAL_PROJECT_PROPOSAL"
author: "John K. Hancock"
date: "October 29, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
#import the required libraries
library(dplyr)
library(httr)
library(jsonlite)
library(tidyverse)  
library(rvest)    
library(stringr)   
library(rebus)
library(tm)
library(wordcloud)
library(viridisLite)
library(ggplot2)
library(stringr)
library(lubridate)
library(GENEAread)
library(Hmisc)
library(scales)
theme_update(plot.title = element_text(hjust = 0.5))
```


#DATA606 Fall 2018 - FINAL PROJECT
#Pitching and Wins

##I. Introduction

In the game of baseball, there's a heuristic that pitching wins games and matters more than hitting and scoring.  In a blog article titled, Hitting or Pitching,Which wins more games, www.wmbriggs.com/post/120/, by Tim Murray and William Briggs (April 28, 2008), they explore which is a better predictor of wins, hitting or pitching.  For this project, I took a slightly different appoach. I wanted to see whether pitching alone is a good predictor of wins.  


##II. Data

The data was collected from the website, fangraphs.com.  Fangraphs is operated by FanGraphs, Inc. Fangraphs compiles historical statistical data for the entire history of Major League Baseball ("MLB"). In addition, it creates and records advanced baseball metrics outside of the established statistics. FanGraphs is well established as a chronicler and compiler of baseball statistics. It has parternership deals with ESPN and SB Nation. Fangraphs has an interface for members (of which I am one) where you are able to download individual and team statistics covering a single season or multiple years. For this project, I downloaded data for all 30 teams from 1998 to 2017.  I purposely chose the year 1998 as that was the last year that MLB expanded its number of teams.On the site, you can generate a custom report for pitching statistics which is how I collected the data. 

I combined the downloads into one csv file which is located in the DATA folder.  The file is titled: DATA/MLB_PITCHING_STATS_1998_to_2017.csv

##III. Exploratory Data Analysis

```{r}
mlb_df <- read.csv('DATA/MLB_PITCHING_STATS_1998_to_2017.csv')
```

```{r}
dim(mlb_df)
```

There are 600 observations and 22 variables. Each observation represents the win total and key pitching statistics for 30 MLB teams from 1998 to 2017. 

###Description of Variables

The table below provides descriptions for all 22 variables. 

Variables | Description
------------- | -------------
Team|Name of the MLB Team
W|Number of Wins
ERA|Team Earned Run Average
H|Number of Hits Allowed by Team
R|Number of Runs Allowed by Team
ER|Earned Runs Allowed by Team
HR|Home Runs Allowed by Team
BB|Walks Allowed by Team
SO|Strikeouts by Team
KsPer9|Number of Strikeouts per 9 innings
BBper9|Number of Walks per 9 innings
KperBB|Strikeouts per Walks
Hper9|Hits allowed per 9 innings
HRper9|Home Runs per 9 innings
AVG|Batting average allowed by Team
WHIP|Walks plus Hits per innings pitched
BABIP|The rate at which the pitcher allows a hit when the ball is put in play, calculated as (H-HR)/(AB-K-HR+SF)
LOB_pct|Percentage of pitcher?s own base runners that they strand over the course of a season Not equal to the LOB column in the box score
FIP|An estimate of a pitcher?s ERA based on strikeouts, walks/HBP, and home runs allowed, assuming league average results on balls in play
RAR|Runs above replacement level The number of runs a player has been worth to his team compared to a freely available player based on FIP
K_pct|Frequency with which the pitcher has struck out a batter, calculated as strikeouts divided by total batters faced
BB_pct|Frequency with which the pitcher has issued a walk, calculated as walks divided by total batters faced 

###Scaling the Data

```{r}
head(mlb_df, 10)
```

Looking at the first 10 records from the load, you can see that the numerical data is on different scales. For example, Hits allowed ("H") is an integer value, but K_pct is a percentage. From the MiniTab blog post, "When Is It Crucial to Standardize the Variables in a Regression Model?" (02/10/2016):

"You should standardize the variables when your regression model contains polynomial terms or interaction terms. While these types of terms can provide extremely important information about the relationship between the response and predictor variables, they also produce excessive amounts of multicollinearity.

Multicollinearity is a problem because it can hide statistically significant terms, cause the coefficients to switch signs, and make it more difficult to specify the correct model.

Your regression model almost certainly has an excessive amount of multicollinearity if it contains polynomial or interaction terms. Fortunately, standardizing the predictors is an easy way to reduce multicollinearity and the associated problems that are caused by these higher-order terms. If you don't standardize the variables when your model contains these types of terms, you are at risk of both missing statistically significant results and producing misleading results."

(MiniTab)[http://blog.minitab.com/blog/adventures-in-statistics-2/when-is-it-crucial-to-standardize-the-variables-in-a-regression-model]

For this project, I re-scaled the data by applying the 




```{r}
#mlb_df[,3:22] <- t(apply(mlb_df[,3:22], 1, rescale))
#head(mlb_df,10)
```

### Creating Categorical Varibles

### Record Category

Looking at the distribution for wins, you can see that Wins are normally distributed. 

```{r}
ggplot(mlb_df, aes(x=W)) + geom_histogram(bins=20,
                 col="blue", 
                 fill="orange") + labs(title="Histogram for Wins", align="center") +
                                  labs(x="Wins", y="Count")
```

For a MLB team, winning 100 games in a season is a high water mark. It is considered a high accomplishment. Over the past 20 years prior to 2018, only 24 times (4%) has a team won 100 games or more. 

```{r}
one_hundred_wins <- nrow(mlb_df[mlb_df$W >= 100,])
one_hundred_wins
one_hundred_wins / nrow(mlb_df)

```

I created a new variable called "Record" which catgorizes the win totals by levels

1. 100 wins or more
2. Above .500 and Less than 100 wins
3. Below .500 but above 50 wins
4. Less than 50 wins

```{r}
mlb_df[mlb_df$W >= 100, "Record"] <- "100 wins or Above"
mlb_df[mlb_df$W >= 82  & mlb_df$W < 100, "Record"] <- "Above .500 and Less than 100 Wins"
mlb_df[mlb_df$W < 82 & mlb_df$W > 50, "Record"] <- "Below .500 but above 50 wins"
mlb_df[mlb_df$W <= 50, "Record"] <- "Less than 50 wins"
mlb_df$Record <- as.factor(mlb_df$Record)

```


### League Category

MLB is an unusual sport in that it consists of two different leagues, the American and National leagues ("AL" and "NL"), that play under different rules.  In the National league, the pitcher, typically the worst hitter, has to bat.  In the American league, teams are allowed a Designated Hitter. 

In short, one league has the potential for more offense, AL, than the other, and this impacts pitching as NL pitchers typically have lower ERAs. For the model, I have to distinguish between the two leagues.  To do so, I created the category variable, "League", and assigned the corresponding league to each team.  


```{r}
mlb_df$League <- case_when(mlb_df$Team == 'Braves'~'National League',
 mlb_df$Team == 'Marlins'~'National League',
 mlb_df$Team == 'Mets'~'National League',
 mlb_df$Team == 'Phillies'~'National League',
 mlb_df$Team == 'Nationals'~'National League',
 mlb_df$Team == 'Cubs'~'National League',
 mlb_df$Team == 'Reds'~'National League',
 mlb_df$Team == 'Brewers'~'National League',
 mlb_df$Team == 'Pirates'~'National League',
 mlb_df$Team == 'Cardinals'~'National League',
 mlb_df$Team == 'iamondbacks'~'National League',
 mlb_df$Team == 'Rockies'~'National League',
 mlb_df$Team == 'Dodgers'~'National League',
 mlb_df$Team == 'Padres'~'National League',
 mlb_df$Team == 'Giants'~'National League',
 TRUE ~ as.character('American League'))
```

During the past 20 years, one team, the Houston Astros changed leagues from the NL to the AL, so I have to account for this switch.


```{r}
rows <- as.numeric(row.names.data.frame(mlb_df[mlb_df$Team == 'Astros',])) 
rows <- rows[rows < 480]
mlb_df$League[ rows] <- "National League"
```


```{r}
mlb_df$League <- as.factor(mlb_df$League)
head(mlb_df[mlb_df$Team == 'Astros', ],5)
tail(mlb_df[mlb_df$Team == 'Astros', ],5)
```

### Predictor Variables


In MLB, there are traditional pitching statistics and more advanced statistics.  


```{r}
describe(mlb_df$ERA)
```

```{r}
ggplot(mlb_df, aes(x=ERA)) + geom_histogram(bins=20,
                 col="white", 
                 fill="red") + labs(title="Histogram for ERA") +
                                  labs(x="Earned Run Average", y="Count")
```











```{r}
describe(mlb_df$SO)
```
```{r}
describe(mlb_df$WHIP)
```


```{r}
describe(mlb_df$FIP)
```

```{r}
table(mlb_df$Record, useNA='ifany')
```



```{r}

```


```{r}
ggplot(mlb_df, aes(x=FIP)) + geom_histogram(bins=15,
                 col="blue", 
                 fill="orange") + labs(title="Histogram for Fielding Independent Pitching") +
                                    labs(x="FIP", y="Count")
```

