---
title: "Chapter 3 Homework"
author: "John K. Hancock"
date: "September 16, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library('DATA606')
library('data.table')
```

####Chapter 3 Homework

##3.2 Area under the curve, Part II. What percent of a standard normal distribution
N(μ =0,! = 1) is found in each region? Be sure to draw a graph.

(a) Z > −1.13 

The area under the curve where the Z score is greater than -1.13 is 0.87. Using the complement Probability 1-P. 
```{r}
1-0.1292
```



```{r}
normalPlot(mean = 0, sd = 1, bounds = c(-1.13, 5), tails = FALSE)

```
(b) Z < 0.18
The area under the curve where the Z score is less than 0.18 is .5714
```{r}
normalPlot(mean = 0, sd = 1, bounds = c(-5, 0.18), tails = FALSE)
```


(c) Z > 8
The area under the curve where the Z score is greater than 8 is 0.  Since all Z greater than or equal to 3.50, the probability is greater than or equal to 0.9998, and using the complement rule, 1-1= 0.
```{r}
normalPlot(mean = 0, sd = 1, bounds = c(0, 0), tails = FALSE)
```

(d) |Z| < 0.5

Can be re-written as -.5 < z > .5 =  P(Z<.5)-P(Z>-.5)  The area under the curve is 0.383

```{r}
0.6915-0.3085
```
```{r}
normalPlot(mean = 0, sd = 1, bounds = c(-.5, .5), tails = FALSE)
```
##3.4 Triathlon times, Part I. In triathlons, it is common for racers to be placed into age
and gender groups. Friends Leo and Mary both completed the Hermosa Beach Triathlon, where
Leo competed in the Men, Ages 30 - 34 group while Mary competed in the Women, Ages 25 -
29 group. Leo completed the race in 1:22:28 (4948 seconds), while Mary completed the race in
1:31:53 (5513 seconds). Obviously Leo finished faster, but they are curious about how they did
within their respective groups. Can you help them? Here is some information on the performance
of their groups:

• The finishing times of the Men, Ages 30 - 34 group has a mean of 4313 seconds with a
standard deviation of 583 seconds.

• The finishing times of the Women, Ages 25 - 29 group has a mean of 5261 seconds with a
standard deviation of 807 seconds.

• The distributions of finishing times for both groups are approximately Normal.

Remember: a better performance corresponds to a faster finish.
(a) Write down the short-hand for these two normal distributions.
###Answer
For men, N(u=4313, sd=583)
For women, N(u=5261, sd=807)

(b) What are the Z-scores for Leo’s and Mary’s finishing times? What do these Z-scores tell you?
###Answer
Leo's z-score is 1.09
```{r}
(4948-4313) / 583
```
Mary's z-score is 0.312
```{r}
(5513-5261) / 807
```
The z-scores show the number of standard deviations either above or below the mean. 


(c) Did Leo or Mary rank better in their respective groups? Explain your reasoning.
###Answer
We see that Leo finished better than Mary did in his group than she did within her's. Leo was more than a full standard deviation above the mean while Mary was only 1/3 above the mean.

(d) What percent of the triathletes did Leo finish faster than in his group?
###Answer
Leo finished faster than 86.21% of the runners while 

(e) What percent of the triathletes did Mary finish faster than in her group?
###Answer
Mary finished faster than 62.17% of the runners in her group.

(f) If the distributions of finishing times are not nearly normal, would your answers to parts
(b) - (e) change? Explain your reasoning.
Yes. If the data was not nearly normal, then the deviations from the mean would

##3.18 Heights of female college students. Below are heights of 25 female college students.
(a) The mean height is 61.52 inches with a standard deviation of 4.58 inches. Use this information
to determine if the heights approximately follow the 68-95-99.7% Rule.
###Answer
Yes, the below calculations show that the data follows the 68-95-99.7% Rule or the Empirical Rule. 

```{r}
data =c(54,55,56,56,57,58,58,59,60,60,60,61,61,62,62,63,63,63,64,65,65,67,67,69,73)


one_sd<-data[between(data,(mean(data)-sd(data)), (mean(data)+sd(data)))]
length(one_sd) / length(data)	 

```


```{r}
two_sd<-data[between(data,(mean(data)-2*sd(data)), (mean(data)+2*sd(data)))]
length(two_sd) / length(data)
```
```{r}
three_sd<-data[between(data,(mean(data)-3*sd(data)), (mean(data)+3*sd(data)))]
length(three_sd) / length(data)
```


(b) Do these data appear to follow a normal distribution? Explain your reasoning using the graphs
provided below.

###Answer
Yes. The grpahs show a unimodal distribution with the highest bar being in the middle of the graph.



