---
title: "Chapter 4 Assignment"
author: "John K. Hancock"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###4.4 Heights of adults. Researchers studying anthropometry collected body girth measurements
###and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically
###active individuals. The histogram below shows the sample distribution of heights in centimeters.

####Answers:

(a) The mean is 171.1, and the median is 170.3

(b) The point estimate for the standard deviation is 9.4, and the IQR is 7.5

(c) Is a person who is 1m 80cm (180 cm) tall considered unusually tall? 

No, the z-value of the observation at 180cm is approx. 1 standard deviation (0.95) above the mean and within the 95% confidence interval. 

And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

No, the z-value of the observation at 155cm is -1.71 standard deviations below the mean and within the 95% confidence interval. 


```{r}
#Observation of 180cm
sample_mean <- 171.1
observation <- 180
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 

```

```{r}
#Observation of 155cm
sample_mean <- 171.1
observation <- 155
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 
```

(d) The researchers take another random sample of physically active individuals. Would you
expect the mean and the standard deviation of this new sample to be the ones given above?
Explain your reasoning.

No. There are variations among different sample populations 


(e) The sample means obtained are point estimates for the mean height of all active individuals,
if the sample of individuals is equivalent to a simple random sample. What measure do we
use to quantify the variability of such an estimate (Hint: recall that SD¯x = ! pn )? Compute
this quantity using the data from the original sample under the condition that the data are a
simple random sample.

To compute the Standard Error, take the Standard Deviation divided by the square root of the sample size which is 0.42.


```{r}
stand_error / sqrt(507)
```

