---
title: "Chapter 4 Assignment"
author: "John K. Hancock"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###4.4 Heights of adults. Researchers studying anthropometry collected body girth measurements
###and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically
###active individuals. The histogram below shows the sample distribution of heights in centimeters.

####Answers:

(a) The mean is 171.1, and the median is 170.3

(b) The point estimate for the standard deviation is 9.4, and the IQR is 7.5

(c) Is a person who is 1m 80cm (180 cm) tall considered unusually tall? 

No, the z-value of the observation at 180cm is approx. 1 standard deviation (0.95) above the mean and within the 95% confidence interval. 

And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

No, the z-value of the observation at 155cm is -1.71 standard deviations below the mean and within the 95% confidence interval. 


```{r}
#Observation of 180cm
sample_mean <- 171.1
observation <- 180
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 

```

```{r}
#Observation of 155cm
sample_mean <- 171.1
observation <- 155
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 
```

(d) The researchers take another random sample of physically active individuals. Would you
expect the mean and the standard deviation of this new sample to be the ones given above?
Explain your reasoning.

No. There are variations among different sample populations 


(e) The sample means obtained are point estimates for the mean height of all active individuals,
if the sample of individuals is equivalent to a simple random sample. What measure do we
use to quantify the variability of such an estimate (Hint: recall that SD¯x = ! pn )? Compute
this quantity using the data from the original sample under the condition that the data are a
simple random sample.

To compute the Standard Error, take the Standard Deviation divided by the square root of the sample size which is 0.42.


```{r}
stand_error / sqrt(507)
```
4.14 Thanksgiving spending, Part I. The 2009 holiday retail season, which kicked o??? on
November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported
consumer spending than was seen during the comparable period in 2008. To get an estimate
of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer
spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber
Monday, averaged $84.71. A 95% confidence interval based on this sample is ($80.31, $89.11).
Determine whether the following statements are true or false, and explain your reasoning.

(a) We are 95% confident that the average spending of these 436 American adults is between
$80.31 and $89.11.

FALSE. We are 95% confident that the average spending for the population is within the range of ($80.31, $89.11). The sample average is a point estimate for the population.

(b) This confidence interval is not valid since the distribution of spending in the sample is right
skewed.

FALSE. Given that the sample was chosen at random and that the sample size is greater than 30, we can apply the Central Limit Thereom to assume normality of the samples. 

(c) 95% of random samples have a sample mean between $80.31 and $89.11.
FALSE. The confidence interval is an estimate for the population not sample sizes. 


(d) We are 95% confident that the average spending of all American adults is between $80.31 and
$89.11.
(e) A 90% confidence interval would be narrower than the 95% confidence interval since we don't
need to be as sure about our estimate.
(f) In order to decrease the margin of error of a 95% confidence interval to a third of what it is
now, we would need to use a sample 3 times larger.
(g) The margin of error is 4.4.


