---
title: "Chapter 4 Assignment"
author: "John K. Hancock"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###4.4 Heights of adults. Researchers studying anthropometry collected body girth measurements
###and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically
###active individuals. The histogram below shows the sample distribution of heights in centimeters.

####Answers:

(a) The mean is 171.1, and the median is 170.3

(b) The point estimate for the standard deviation is 9.4, and the IQR is 7.5

(c) Is a person who is 1m 80cm (180 cm) tall considered unusually tall? 

No, the z-value of the observation at 180cm is approx. 1 standard deviation (0.95) above the mean and within the 95% confidence interval. 

And is a person who is 1m 55cm (155cm) considered unusually short? Explain your reasoning.

No, the z-value of the observation at 155cm is -1.71 standard deviations below the mean and within the 95% confidence interval. 


```{r}
#Observation of 180cm
sample_mean <- 171.1
observation <- 180
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 

```

```{r}
#Observation of 155cm
sample_mean <- 171.1
observation <- 155
stand_error <- 9.4 

(observation - sample_mean)/ stand_error 
```

(d) The researchers take another random sample of physically active individuals. Would you
expect the mean and the standard deviation of this new sample to be the ones given above?
Explain your reasoning.

No. There are variations among different sample populations 


(e) The sample means obtained are point estimates for the mean height of all active individuals,
if the sample of individuals is equivalent to a simple random sample. What measure do we
use to quantify the variability of such an estimate (Hint: recall that SD¯x = ! pn )? Compute
this quantity using the data from the original sample under the condition that the data are a
simple random sample.

To compute the Standard Error, take the Standard Deviation divided by the square root of the sample size which is 0.42.


```{r}
stand_error / sqrt(507)
```
4.14 Thanksgiving spending, Part I. The 2009 holiday retail season, which kicked o??? on
November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported
consumer spending than was seen during the comparable period in 2008. To get an estimate
of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer
spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber
Monday, averaged $84.71. A 95% confidence interval based on this sample is ($80.31, $89.11).
Determine whether the following statements are true or false, and explain your reasoning.

(a) We are 95% confident that the average spending of these 436 American adults is between
$80.31 and $89.11.

FALSE. We are 95% confident that the average spending for the population is within the range of ($80.31, $89.11). The sample average is a point estimate for the population.

(b) This confidence interval is not valid since the distribution of spending in the sample is right
skewed.

FALSE. Given that the sample was chosen at random and that the sample size is greater than 30, we can apply the Central Limit Thereom to assume normality of the samples. 

(c) 95% of random samples have a sample mean between $80.31 and $89.11.
FALSE. The confidence interval is an estimate for the population not sample sizes. 


(d) We are 95% confident that the average spending of all American adults is between $80.31 and
$89.11.

TRUE. The confidence interval shows that the true average spending for all Americans is between $80.31 and
$89.11. 

(e) A 90% confidence interval would be narrower than the 95% confidence interval since we don't
need to be as sure about our estimate.

TRUE. At 90% confidence interval, the range would be narrower. 

(f) In order to decrease the margin of error of a 95% confidence interval to a third of what it is
now, we would need to use a sample 3 times larger.

FALSE. In order to reduce the Standard Error by a third, the formula requires that the sample size be nine times the oringal sample since the formula is: sample mean * z-score * sqrt(sample size)


(g) The margin of error is the confidence interval divided by 2.

```{r}
(89.11-80.31) / 2

```

4.24 Gifted children, Part I. Researchers investigating characteristics of gifted children collected
data from schools in a large city on a random sample of thirty-six children who were identified
as gifted children soon after they reached the age of four. The following histogram shows the distribution
of the ages (in months) at which these children first counted to 10 successfully. Also
provided are some sample statistics.43

(a) Are conditions for inference satisfied?

Yes. Sample was chosen randomly.  The sample is less than 10% of the population which means that it's indepdendent, and the sample size is greater than 30 which can be approximated to be normal given the Central Limit Theorem.


(b) Suppose you read online that children first count to 10 successfully when they are 32 months
old, on average. Perform a hypothesis test to evaluate if these data provide convincing evidence
that the average age at which gifted children fist count to 10 successfully is less than the general
average of 32 months. Use a significance level of 0.10.


Null Hypothesis: Students count to 10 when they're 32 months old. 
Alternative Hypothesis: Students count to 10 when they're less than 32 months old. 

```{r}
sample_mean2 <- 30.69
sample_std2 <- 4.31
sample_size2 <- 36
population_mean2 <- 32
stand_error2 <- (sample_std2) /sqrt(sample_size2)

(sample_mean2 - population_mean2) / (sample_std2 / sqrt(sample_size2))





```

The z-score is -1.824 which corresponds to 0.9656.  The p value is 0.0344 which is less than .10.  Therefore, we can reject the null hypothesis. 

```{r}
1-0.9656
```


(c) Interpret the p-value in context of the hypothesis test and the data.

The p-value is the probability of observing data at least as favorable to the alternative
hypothesis as our current data set. If the null hypothesis is true, the probability of observing a sample mean less than 32 months is only 3.4%


(d) Calculate a 90% confidence interval for the average age at which gifted children first count to
10 successfully.


```{r}
error <- 1.65 * stand_error2 
left<- sample_mean2-error
right <- sample_mean2+error

left
right

```
A 90% Confidence Interval would be (29.50475, 31.87525)


(e) Do your results from the hypothesis test and the confidence interval agree? Explain.

Yes. The sample mean is 32 months, and although it's at the border of the upper bound border for the 90% Confidence Interval, is not within the interval.. This along with the hypothesis test which rejected the null hypothesis both agree. 



4.26 Gifted children, Part II. Exercise 4.24 describes a study on gifted children. In this study,
along with variables on the children, the researchers also collected data on the mother's and father's
IQ of the 36 randomly sampled gifted children. The histogram below shows the distribution of
mother's IQ. Also provided are some sample statistics.


(a) Perform a hypothesis test to evaluate if these data provide convincing evidence that the average
IQ of mothers of gifted children is di???erent than the average IQ for the population at large,
which is 100. Use a significance level of 0.10.

The Null Hypothesis is: IQ of Mothers of Gifted children  = 100
The alternative hypothesis is: IQ of Mothers of Gifted children  != 100

The z-score for the sample mean is 16.8.  The p-score is the probability of getting that z-score < -16.8 is 0 which is less than .10. Thus, We can reject the null hypothesis


```{r}
sample_mean3 <- 118.2
sample_std3 <- 6.5
sample_size3 <- 36
population_mean3 <- 100


(sample_mean3 - population_mean3) / ((sample_std3) / sqrt(sample_size3))



```


```{r}
stand_error3 <- (sample_std3) / sqrt(sample_size3)


error2 <- 1.65 * stand_error3

left2 <- sample_mean3 - error2
right2 <- sample_mean3 + error2

left2
right2
```

(b) Calculate a 90% confidence interval for the average IQ of mothers of gifted children.

The 90% Confidence Interval for the sample population is 116.4125 to 119.9875.


(c) The entire 90% confidence interval for the Mothers of Gifted children is above the mean IQ for the entire population and given the results of the p-test, both results agree.


4.34 CLT. Define the term "sampling distribution" of the mean, and describe how the shape,
center, and spread of the sampling distribution of the mean change as sample size increases.


