---
title: "Chapter 4 Assignment"
author: "John K. Hancock"
date: "September 25, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

###4.4 Heights of adults. Researchers studying anthropometry collected body girth measurements
###and skeletal diameter measurements, as well as age, weight, height and gender, for 507 physically
###active individuals. The histogram below shows the sample distribution of heights in centimeters.

Answers: (a) The mean is 171.1 cm, and the median is 170.3 cm
         (b) The point estimate for the standard deviation is 9.4 cm, and the IQR is 7.5 cm
         (c) Neither the person that's 180 cm tall or the person that's 55 cm tall is not unusual because it's within 2 positive/negative standard deviations of the mean. 
         (d) No. There will be variations for the mean and standard deviation of the next sample taken than the original. The current sample is just an estimate.
         (e) We would use the Standard Error to quantify the variability of the mean of this sample. (.417)
         
```{r}
#Calculations for (c)
print((180 - 171.1) / 9.4)
print((155-171.1) / 9.4)
```
         
```{r}
#Calculations for (e)
print(9.4 / (sqrt(507)))

```

###4.14 Thanksgiving spending, Part I. The 2009 holiday retail season, which kicked o??? on
###November 27, 2009 (the day after Thanksgiving), had been marked by somewhat lower self-reported
###consumer spending than was seen during the comparable period in 2008. To get an estimate
###of consumer spending, 436 randomly sampled American adults were surveyed. Daily consumer
###spending for the six-day period after Thanksgiving, spanning the Black Friday weekend and Cyber
###Monday, averaged $84.71. A 95% confidence interval based on this sample is ($80.31, $89.11).
###Determine whether the following statements are true or false, and explain your reasoning.         

(a) We are 95% confident that the average spending of these 436 American adults is between
$80.31 and $89.11.

Answer: False. We are 95% confident that the true population mean for spending is within the range $80.31 and $89.11.

(b) This confidence interval is not valid since the distribution of spending in the sample is right
skewed.

Answer: False. As the sample size increases, the normal model for the sample mean becomes more reasonable. We
can also relax our condition on skew when the sample size is very large. In this problem, the sample size is 436, greater than 30.

(c) 95% of random samples have a sample mean between $80.31 and $89.11.
Answer: False. The 95% CI is an estimate for the population mean, not the sample mean.

(d) We are 95% confident that the average spending of all American adults is between $80.31 and
$89.11.

Answer: True. The 95% CI is an estimate for the population mean. We are 95% confident that the population mean is between $80.31 and
$89.11.

(e) A 90% confidence interval would be narrower than the 95% confidence interval since we don't
need to be as sure about our estimate.

Answer: True. A 90% CI is a narrower estimate for the population mean. We would be only are 90% confident for a range for the population mean.

(f) In order to decrease the margin of error of a 95% confidence interval to a third of what it is
now, we would need to use a sample 3 times larger.

Answer: False. To reduce the margin of error by a third, you would need a sample size nine times larger. 

(g) The margin of error is 4.4.

Answer: True.

```{r}
print(84.71-4.4)

print(84.71+4.4)


```
###4.24 Gifted children, Part I. Researchers investigating characteristics of gifted children collected
###data from schools in a large city on a random sample of thirty-six children who were identified
###as gifted children soon after they reached the age of four. The following histogram shows the distribution
###of the ages (in months) at which these children first counted to 10 successfully. Also
###provided are some sample statistics.43

n= 36
min= 21
mean= 30.69
sd= 4.31
max= 39

(a) Are conditions for inference satisfied?

Sample is less than 10% of the population, greater than 30, chosen randomly, and being over 30 means that the sample follows a normal distribution.  Yes. 

(b) Suppose you read online that children first count to 10 successfully when they are 32 months
old, on average. Perform a hypothesis test to evaluate if these data provide convincing evidence
that the average age at which gifted children first count to 10 successfully is less than the
general average of 32 months. Use a significance level of 0.10.

Null Hypothesis: The average age of gifted children to first count to 10 is >=  32 mos.
Alternative Hypothesis: The average age of gifted children to first count to 10 is < 32 mos.



